---
title: 'Excel to R: Part Two'
author: "Adnan Shroufi"
date: "2022-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
options(scipen = 999)
```

# Introduction -----------------------------------------------------------------

## Introduction to Excel -> R: Part One

This is a follow-up to the first Excel to R C&C session.
That session focused mostly on using the {dplyr} and {tidyr} packages, for data processing tasks like those you'd do in Excel.
As mentioned in the first session, while Excel is the appropriate tool for some tasks, it does have some limitations, which include:

- Excel can be slow when working with large files
- Going over the same steps in Excel every day/week/month to generate the same report/analysis is time consuming and repetitive.

Using R can can overcome these issues for the following reasons:

- R can work with large files far more easily and quickly
- A single R script can do all of the time consuming and repetitive tasks for you

## Session Format

This session will look at a real example of an excel workflow that has been turned into an R script.
The uses various time and date fields in order to generate some 

## Load Packages

Before we load in our data, we need to make sure have all the required packages installed.
A 'package' is basically a collection of functions. You need to install a package first before you can use it.
We can use `install.package("package_name")` to install a package.
We then use `library(package_name)` to load the package.

There are 2 packages we'll be using quite a lot.
Those are {dplyr} and {lubridate}, for data processing and working with dates.
The other six packages will be used infrequently.

```{r, echo=FALSE}
# Frequently used libraries 
library(dplyr)
library(lubridate)

# Infrequently used libraries 
library(openxlsx)
library(bizdays)
library(readr)
library(scales)
library(jsonlite)
library(janitor)
library(stringr)
```

## Working with dates in R

Many typical reporting tasks in Excel use date fields.
Numerical or character formatted columns can be converted to a date format in R.
Once in a date format, we can then easily do all sorts of date-related calculations.

First we will load in some dummy dates with some unformatted dates.
We will also use the str() function to check the 'structure' of this data, i.e. what data types the columns are.
Each columns is a character (chr).

```{r, echo=FALSE}
# Load in dummy dates data
dummy_dates = read.csv("dummy_dates.csv")

# Check structure of the initial data
str(dummy_dates)
```

## Format each date field accordingly

We can change a character to a date using as.Date() within a mutate().
mutate() either creates a new column or overwrites an existing column.
The only difficulty here is specifying the format.
The format depends on how the character has been spelt.

We have to tell R whether to look for a '/', '-', or '.' between the date elements.
We then have to tell it what position the year (Y), month (m) and (d) come.

```{r, echo=FALSE}
dummy_dates = dummy_dates %>% 
  mutate(date_one = as.Date(date_one, format = "%d/%m/%Y")) %>% 
  mutate(date_two = as.Date(date_two, format = "%d.%m.%Y")) %>% 
  mutate(date_three = as.Date(date_three, format = "%Y-%d-%m")) %>% 
  mutate(date_four = as.Date(date_four, format = "%d/%m/%Y"))

# Re-check structure of the now processed data
str(dummy_dates)
```

## Some very simple examples of date field calculations

Once in a date format, various calculations can be easily done.
Note, if your analysis needs 

```{r, echo=FALSE}
dummy_dates %>% 
  mutate(
    diff = date_one - date_two,
    day = lubridate::day(date_one),
    month = lubridate::month(date_one),
    year = lubridate::year(date_one)
  )
```



```{r, echo=FALSE}

```


```{r, echo=FALSE}

```


```{r, echo=FALSE}

```


```{r, echo=FALSE}

```


```{r, echo=FALSE}

```


```{r, echo=FALSE}

```


```{r, echo=FALSE}

```


```{r, echo=FALSE}

```


```{r, echo=FALSE}

```

# Define file path so don't need to repeat it
file_path = "C:/Users/adnsh/OneDrive - NHS Business Services Authority/Desktop/Excel to R/Wait_Times_GHIC_MAR_2022_ZIP.csv"

# check structure of data if loaded with base-R
str(read.csv(file_path))

# Check structure of data if loaded with 'readr'
str(readr::read_csv(file_path))

# Have a look at some rows of the data 
head(data)

# Create report date
report_date <- as.Date(lubridate::today())

# Process data
data_edit = data %>% 
  # Clean names
  janitor::clean_names() %>% 
  # Rename one variable
  rename(submitted_date = submitted_to_xerox) %>% 
  # change all posixct to date format
  mutate(application_date = as.Date(application_date)) %>% 
  mutate(issued_date = as.Date(issued_date)) %>% 
  mutate(expired_date = as.Date(expired_date)) %>% 
  mutate(rejected_date = as.Date(rejected_date)) %>% 
  # Create a 'start date' field 
  mutate(
    start_date = case_when(
      !is.na(rejected_date) ~ rejected_date,
      !is.na(reissue_request_date) ~ reissue_request_date,
      !is.na(issued_date) ~ issued_date,
      T ~ as.Date("3099-12-31")
    )
  ) %>% 
  # Create a 'end date' field
  mutate(
    end_date = case_when(
      !is.na(rejected_date) ~ rejected_date,
      !is.na(dispatched_date) ~ dispatched_date,
      T ~ report_date
    )
  ) %>% 
  # Create a 'status' field
  mutate(
    status = case_when(
      start_date == "3099-12-31" ~ "MISSING START DATE",
      !is.na(rejected_date) ~ "REJECTED",
      !is.na(dispatched_date) ~ "DISPATCHED",
      !is.na(submitted_date) ~ "SUBMITTED",
      !is.na(issued_date) ~ "ISSUED",
      !is.na(reissue_request_date) ~ "REREQUEST",
      T ~ "ERROR"
    )
  ) %>% 
  # Create a 'dispatch month' field
  mutate(dispatch_month = floor_date(dispatched_date, unit = "month"))

# Newly edited data
data_coalesce = data %>% 
  # Clean names
  janitor::clean_names() %>% 
  # Rename one variable
  rename(submitted_date = submitted_to_xerox) %>% 
  # change all posixct to date format
  mutate(application_date = as.Date(application_date)) %>% 
  mutate(issued_date = as.Date(issued_date)) %>% 
  mutate(expired_date = as.Date(expired_date)) %>% 
  mutate(rejected_date = as.Date(rejected_date)) %>% 
  # Create a 'start date' and 'end date' field 
  mutate(start_date = coalesce(rejected_date, reissue_request_date, issued_date, as.Date("3099-12-31"))) %>% 
  mutate(end_date = coalesce(rejected_date, dispatched_date, report_date)) %>% 
  # Create a 'status' field
  mutate(
    status = case_when(
      start_date == "3099-12-31" ~ "MISSING START DATE",
      !is.na(rejected_date) ~ "REJECTED",
      !is.na(dispatched_date) ~ "DISPATCHED",
      !is.na(submitted_date) ~ "SUBMITTED",
      !is.na(issued_date) ~ "ISSUED",
      !is.na(reissue_request_date) ~ "REREQUEST",
      T ~ "ERROR"
    )
  ) %>% 
  # Create a 'dispatch month' field
  mutate(dispatch_month = floor_date(dispatched_date, unit = "month"))

# check if outputs are the same
setdiff(data_edit, data_coalesce)

# Retrieve and format json data
dates = jsonlite::read_json("https://www.gov.uk/bank-holidays.json")

# convert date info to dataframe with a single column
holiday_dates = dates$`england-and-wales`$events %>% 
  bind_rows() %>% 
  select(date) %>% 
  rename(Date = date) %>% 
  as.data.frame() %>% 
  mutate(Date = as.Date(Date))

# Check data type of dates
str(holiday_dates)

#Working with bizdays to calculate time
my_calendar = bizdays::create.calendar(
  name = "WorkCal",
  holidays = holiday_dates$Date,
  weekdays = c("saturday", "sunday"),
  start.date = as.Date("2020-01-01"),
  end.date = as.Date("2023-12-31")
  )

# check holidays within the calendar
my_calendar$holidays

# Final data
data_edit = data_edit %>% 
  mutate(sla_days = bizdays::bizdays(start_date, end_date, my_calendar)) %>% 
  mutate(in_sla = ifelse(sla_days <= 10, "Yes", "No")) %>% 
  filter(status == "DISPATCHED")

# Table
table(data_edit$status)
table(data_coalesce$status)

# Remove object and clean
rm(data, data_coalesce)
gc()

# Format final data
output = data_edit %>% 
  mutate(type = "GHIC") %>% 
  group_by(type, dispatch_month, in_sla) %>% 
  summarise(count_by_sla = n()) %>% 
  ungroup() %>% 
  mutate(total = sum(count_by_sla)) %>% 
  mutate(percent = scales::percent(count_by_sla / total, accuracy = 0.00001)) %>% 
  rename_all(.funs = stringr::str_to_title) %>% 
  mutate(Dispatch_month = paste0(
    lubridate::month(Dispatch_month, label = T), " - ", lubridate::year(Dispatch_month)
  ))

# Inspect final data
output

# Check directory
getwd()

# Format output
excel_sheets = list(
  output,
  holiday_dates
)

# Save simple output
openxlsx::write.xlsx(excel_sheets, file = "SIMPLE_OUTPUT.xlsx")

# Create formatted output
wb = createWorkbook()

# Define sheet names
addWorksheet(wb, "March Output")
addWorksheet(wb, "Holiday Dates")

# Modify base font for cells
modifyBaseFont(wb, fontSize = 10, fontName = "Arial")

# Create manual header style
header_style <- createStyle(
  textDecoration = "BOLD", 
  fontColour = "white", 
  fgFill = "#4F81BD",
  border = c("top", "bottom", "left", "right")
)

# Write data to sheets and apply style and specify borders
writeData(wb, "March Output", output, headerStyle = header_style, borders = "columns")
writeData(wb, "Holiday Dates", holiday_dates, headerStyle = header_style, borders = "columns")

# Define column widths
setColWidths(wb, sheet = 1, cols = 1:6, widths = c(10, 18, 10, 15, 10, 10))
setColWidths(wb, sheet = 2, cols = 1, widths = 10)

# Save formatted data
saveWorkbook(wb, "FORMATTED_OUTPUT.xlsx", overwrite = T)

#-------------------------------------------------------------------------------

# Funtions used

# - lubridate::today()
# - lubridate::month()
# - lubridate::year()
# - lubridate::floor_date()

# - dplyr::select()
# - dplyr::mutate()
# - dplyr::case_when()
# - dplyr::filter()
# - dplyr::group_by()
# - dplyr::summmarise()
# - dplyr::ungroup()
# - dplyr::coalesce()
# - dplyr::rename()
# - dplyr::rename_all()
# - dplyr::bind_rows()

# - openxlsx::write.xlsx()
# - openxlsx::createWorkbook()
# - openxlsx::addWorksheet()
# - openxlsx::modifyBaseFont()
# - openxlsx::createStyle()
# - openxlsx::writeData()
# - openxlsx::setColWidths()
# - openxlsx::saveWorkbook()

# - stringr::stringr_to_title()

# - jsonlite::read_json()

# - scales::percent()

# - readr::read_csv()

# - janitor::clean_names()

# - bizdays::bizdays()
# - bizdays::create.calendar()

# - str()
# - setdiff()
# - rm()
# - gc()
# - table()
# - sum()
# - paste0()
# - ifelse()
# - as.Date()
# - list()
# - getwd()

#-------------------------------------------------------------------------------
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
