---
title: "Excel -> R: Part One"
author: "Adnan Shroufi"
date: "10/05/2022"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
options(scipen = 999)
```

# Introduction -----------------------------------------------------------------

## Introduction to Excel -> R: Part One

Lots of people in the NHSBSA use Excel for various kinds of reporting and analysis.
Excel is a great tool with lots of benefits, although has some limitations. These include:

- Excel can be slow when working with large files
- Going over the same steps in Excel every day/week/month to generate the same report/analysis is time consuming and repetitive.

Using R can can overcome these issues for the following reasons:

- R can work with large files far more easily and quickly
- A single R script can do all of the time consuming and repetitive tasks for you

## Load Packages

Before we load in our data, we need to make sure have all the required packages installed.
A 'package' is basically a collection of functions. You need to install a package first before you can use it.
We can use `install.package("package_name")` to install a package.
We then use `library(package_name)` to load the package.

The two packages we are using mostly today are `dplyr` and `tidyr`.
These two packages are often used for data transformation, processing and tidying. 

```{r}
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("stringr")
#install.packages("openxlsx)

library(dplyr)
library(tidyr)
library(stringr)
library(openxlsx)
```

## Load Data

The data we will be using today is a subset of the Open Data Portal (ODP) English Prescribing Data (EPD) data from February 2022.
An EPD dataset for one month is several hundred MBs. The selection we are using today:

- Only uses 9 out of the 26 fields recorded within the EPD dataset
- Only looks at a selection of CCG within the North East Region
- As a result has just over 1m rows, as opposed to ~17m rows
- Has been converted from to CSV to Excel format, for the purposes of today's session 

```{r, echo=FALSE}
data = openxlsx::read.xlsx(xlsxFile = "C:/Users/adnsh/OneDrive - NHS Business Services Authority/Desktop/ODP_EPD_SUBSET_202202.xlsx", sheet = 1)
```

## Format of the Session

We will process and transform the loaded dataset in 4 different ways.
Each of these will demonstrate the use of data processing and transformation functions.
The four processed and transformed datasets will then be collated into a single Excel file and saved.

# Dataset 1 -------------------------------------------------------------------

## Dataset 1.1: Viewing the Data

We can view, filter and inspect the data we have just loaded by clicking on the dataframe name in the global environment.

## Dataset 1.2: Renaming Variables

We cannot manually rename variables in the data viewer, like you could in Excel.
We have to use the rename function to do this.

However, if a new column name has a space it needs to needs to be inside backward ticks.
Or, the words within the column name need to be connected with an underscore.
Purely for readability, and to demonstrate how to format data in R, will we use backward ticks to do this.

```{r}
data %>% 
  # Rename columns
  rename(
    CCG = PCO_NAME,
    `BNF Description` = BNF_DESCRIPTION,
    `BNF Code` = BNF_CODE,
    Quantity = QUANTITY,
    Items = ITEMS
  ) %>% 
  select(CCG, `BNF Description`, `BNF Code`, Quantity, Items)
```
 
## Dataset 1.3: Aggregating the Data

We will now use group_by() to aggregate the data.
We will aggregate by CCG and BNF Description (each of which has a corresponding BNF Code).
The number of distinct categories within each of these fields will determine the size of the resulting dataframe.

Note, the select() function is used to (unsurprisingly) select which columns you want.

To aggregate by CCG and BNF Code, we include these in the group_by(), then decide what function to use within the summarise().
For example, this could be min(), max(), sum(), or such like.
Finally, we must remember to ungroup() the data, otherwise this may affect later calculations.

The following three function can therefore be seen as being used collectively:

- group_by()
- summarise()
- ungroup()

```{r}
mutated_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    `BNF Description` = BNF_DESCRIPTION,
    `BNF Code` = BNF_CODE,
    Quantity = QUANTITY,
    Items = ITEMS
  ) %>% 
  # Aggregate Data
  group_by(CCG, `BNF Description`, `BNF Code`) %>% 
  summarise(
    Items = sum(Items),
    NIC = sum(NIC),
    Quantity = sum(Quantity)
    ) %>% 
  ungroup()
```

## Dataset 1.4: Creating new Columns

The most common way to creating a new column, or edit an existing column, is by using mutate().
Within mutate(), you specify your new coluumn name, plus how you want this column to be created.
If the specified name is that of an existing column, this will merely overwrite this column.

We will create 5 new columns through the following methods:

- Multipying columns
- Dividing columns
- Taking a substring of a column
- Pasting (concatenating) 2 columns together
- Having a static value as a column

```{r}
mutated_data = data %>% 
  # Rename columns
  rename(
    CCG = PCO_NAME,
    `BNF Description` = BNF_DESCRIPTION,
    `BNF Code` = BNF_CODE,
    Quantity = QUANTITY,
    Items = ITEMS
  ) %>% 
  # Aggregate data
  group_by(CCG, `BNF Description`, `BNF Code`) %>% 
  summarise(
    Items = sum(Items),
    NIC = sum(NIC),
    Quantity = sum(Quantity)
    ) %>% 
  ungroup() %>% 
  # Create new columns
  mutate(`Items by Quantity` = Items * Quantity) %>% 
  mutate(`Avg Item Cost` = NIC / Items) %>% 
  mutate(Chapter = substr(`BNF Code`, 1, 2)) %>% 
  mutate(`BNF Description plus Code` = paste0(`BNF Description`, " - ", `BNF Code`)) %>% 
  mutate(`Year Month` = "Feb - 2020")
```

## Dataset 1.5: Editing Existing Columns

We can now edit existing columns, through a column 'mutating itself'.
We will round the Avg Item Cost and replace the "CCG" characters of the CCG column with an empty space.
We can then also trim the CCG column of any white-space, as we have just done some string editing. 

```{r}
mutated_data = data %>% 
  # Rename columns
  rename(
    CCG = PCO_NAME,
    `BNF Description` = BNF_DESCRIPTION,
    `BNF Code` = BNF_CODE,
    Quantity = QUANTITY,
    Items = ITEMS
  ) %>% 
  # Aggregate data
  group_by(CCG, `BNF Description`, `BNF Code`) %>% 
  summarise(
    Items = sum(Items),
    NIC = sum(NIC),
    Quantity = sum(Quantity)
    ) %>% 
  ungroup() %>% 
  # Create new columns
  mutate(`Items by Quantity` = Items * Quantity) %>% 
  mutate(`Avg Item Cost` = NIC / Items) %>% 
  mutate(Chapter = substr(`BNF Code`, 1, 2)) %>% 
  mutate(`BNF Description plus Code` = paste0(`BNF Description`, " - ", `BNF Code`)) %>% 
  mutate(`Year Month` = "Feb - 2020") %>% 
  # Edit existing columns
  mutate(CCG = gsub("CCG", "", CCG)) %>% 
  mutate(CCG = trimws(CCG)) %>% 
  mutate(`Avg Item Cost` = round(`Avg Item Cost`, 2))
```

## Dataset 1.6: Select and arrange columns

You can select/drop columns in two ways.  
By specifying which ones you want to keep.  
We can sometimes use the everything() function to help with this. 
Or specifying which ones you want to drop.

```{r}
mutated_data = data %>% 
  # Rename columns
  rename(
    CCG = PCO_NAME,
    `BNF Description` = BNF_DESCRIPTION,
    `BNF Code` = BNF_CODE,
    Quantity = QUANTITY,
    Items = ITEMS
  ) %>% 
  # Aggregate data
  group_by(CCG, `BNF Description`, `BNF Code`) %>% 
  summarise(
    Items = sum(Items),
    NIC = sum(NIC),
    Quantity = sum(Quantity)
    ) %>% 
  ungroup() %>% 
  # Create new columns
  mutate(`Items by Quantity` = Items * Quantity) %>% 
  mutate(`Avg Item Cost` = NIC / Items) %>% 
  mutate(Chapter = substr(`BNF Code`, 1, 2)) %>% 
  mutate(`BNF Description plus Code` = paste0(`BNF Description`, " - ", `BNF Code`)) %>% 
  mutate(`Year Month` = "Feb - 2020") %>% 
  # Edit existing columns
  mutate(CCG = gsub("CCG", "", CCG)) %>% 
  mutate(CCG = trimws(CCG)) %>% 
  mutate(`Avg Item Cost` = round(`Avg Item Cost`, 2)) %>% 
  # Select columns and arrange values
  select(-c(`BNF Description`, `BNF Code`)) %>% 
  select(`Year Month`, CCG, Chapter, `BNF Description plus Code`, everything()) %>% 
  arrange(CCG, Chapter, `BNF Description plus Code`)
```

## Dataset 1.7: Combining Mutate Commands

If you want, you can perform all of the mutates within a single function, to simplify the code.
Don't worry about this too much if you are still learning about how R works.
We won't be doing this from here on in, so each mutate() step is more clearly visible. 

```{r}
mutated_data = data %>% 
  # Rename columns
  rename(
    CCG = PCO_NAME,
    `BNF Description` = BNF_DESCRIPTION,
    `BNF Code` = BNF_CODE,
    Quantity = QUANTITY,
    Items = ITEMS
  ) %>% 
  # Aggregate data
  group_by(CCG, `BNF Description`, `BNF Code`) %>% 
  summarise(
    Items = sum(Items),
    NIC = sum(NIC),
    Quantity = sum(Quantity)
    ) %>% 
  ungroup() %>% 
  # Create new columns
  mutate(
    `Items by Quantity` = Items * Quantity,
    `Avg Item Cost` = NIC / Items,
    Chapter = substr(`BNF Code`, 1, 2),
    `BNF Description plus Code` = paste0(`BNF Description`, " - ", `BNF Code`),
    `Year Month` = "Feb - 2020",
    # Edit existing columns
    CCG = gsub("CCG", "", CCG),
    CCG = trimws(CCG),
    `Avg Item Cost` = round(`Avg Item Cost`, 2)
    ) %>% 
  # Select columns and arrange values
  select(-c(`BNF Description`, `BNF Code`)) %>% 
  select(`Year Month`, CCG, Chapter, `BNF Description plus Code`, everything()) %>% 
  arrange(CCG, Chapter, `BNF Description plus Code`)

gc()
```

## Dataset 2 -------------------------------------------------------------------

## Dataset 2.1: Create a Chapter Variables and Turn into Numeric

We will first sellect and rename the columns we need, then ccreate a chapter variable.
After that, we will change this from a character (or varchar) to integer.
We can only filter a column numerically, if it is an integer (or numeric) data type.

Also, it is possible to select and rename at the same time if you want, through renaming inside a select().

```{r}
data %>%
  # Rename data
  select(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS
  ) %>% 
  # Create a numeric Chapter variable
  mutate(Chapter = substr(Chapter, 1, 2)) %>% 
  mutate(Chapter = as.integer(Chapter))
```

## Dataset 2.2: Both Character and Numeric Filters

We can use filter() to filter a dataset.
Here is a summary of often used filter commands:

1. == - equals
2. != - does not equal
3. <= - less than or equal to
4. >= - greater than or equal to

Here, we are filtering so we only keep chapters 1-10.
We are filtering 'out' practices labelled as being 'UNIDENTIFIED DOCTORS'.
Also, using grepl(), we are filtering CCG that contain the word 'TYNESIDE'.

```{r}
pivot_long_data = data %>%
  # Rename data
  select(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS
  ) %>% 
  # Create a numeric Chapter variable
  mutate(Chapter = substr(Chapter, 1, 2)) %>% 
  mutate(Chapter = as.integer(Chapter)) %>% 
  # Numeric filter
  filter(Chapter <= 10) %>% 
  # Character filter 
  filter(grepl("TYNESIDE", CCG)) %>% 
  filter(Practice != "UNIDENTIFIED DOCTORS")
```

## Dataset 2.3: Using ifelse and mutate to create a new column

An 'ifelse' command does one of two things, depending on what condition is specified.
Here, any record from chapter 1,2,3,4 or 5 will be described as '1-5' (in a new column).
While any record from chapter 6,7,8,9 or 10 will be described as '6-10' (in the same enw column).

```{r}
pivot_long_data = data %>%
  # Rename data
  select(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS
  ) %>% 
  # Create a numeric Chapter variable
  mutate(Chapter = substr(Chapter, 1, 2)) %>% 
  mutate(Chapter = as.integer(Chapter)) %>% 
  # Numeric filter
  filter(Chapter <= 10) %>% 
  # Character filter 
  filter(grepl("TYNESIDE", CCG)) %>% 
  filter(Practice != "UNIDENTIFIED DOCTORS") %>% 
  # Using ifelse() and mutate()
  mutate(Chapters = ifelse(Chapter <= 5, "1-5", "6-10"))
```

## Dataset 2.4: Group by new column

We can group_by() this new column.
This new column only has 2 values (rather than 10 values).
Note, you can summarise() more than one column within a group().

```{r}
pivot_long_data = data %>%
  # Select and rename data
  select(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS,
    NIC
  ) %>% 
  # Create a numeric Chapter variable
  mutate(Chapter = substr(Chapter, 1, 2)) %>% 
  mutate(Chapter = as.integer(Chapter)) %>% 
  # Numeric filter
  filter(Chapter <= 10) %>% 
  # Character filter 
  filter(grepl("TYNESIDE", CCG)) %>% 
  filter(Practice != "UNIDENTIFIED DOCTORS") %>% 
  # Using ifelse() and mutate()
  mutate(Chapters = ifelse(Chapter <= 5, "1-5", "6-10")) %>% 
  # Group by new 'Chapters' column, to sum() items
  group_by(CCG, Practice, Chapters) %>% 
  summarise(
    Items = sum(Items),
    NIC = sum(NIC)
    ) %>% 
  ungroup()
```

## Dataset 2.5: Pivot data 'long'

We can pivot multiple columns into a single column using pivot_long() from `tidyr`.
We will pivot Items and NIC into a single column.
This will double the number of rows that we have, whilst changing the column format.
You need to specify which columns to 'collapse', how you want the new resultant columns named.

```{r}
pivot_long_data = data %>%
  # Select and rename data
  select(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS,
    NIC
  ) %>% 
  # Create a numeric Chapter variable
  mutate(Chapter = substr(Chapter, 1, 2)) %>% 
  mutate(Chapter = as.integer(Chapter)) %>% 
  # Numeric filter
  filter(Chapter <= 10) %>% 
  # Character filter 
  filter(grepl("TYNESIDE", CCG)) %>% 
  filter(Practice != "UNIDENTIFIED DOCTORS") %>% 
  # Using ifelse() and mutate()
  mutate(Chapters = ifelse(Chapter <= 5, "1-5", "6-10")) %>% 
  # Group by new 'Chapters' column, to sum() items
  group_by(CCG, Practice, Chapters) %>% 
  summarise(
    Items = sum(Items),
    NIC = sum(NIC)
    ) %>% 
  ungroup() %>% 
  # Pivot 'long'
  pivot_longer(cols = c("Items", "NIC"), names_to = "Metric", values_to = "Value")
```

## Dataset 2.6: Format Text Appearance

Finally, we can use str_to_title() from `stringr` to format the appearance of some cells.

```{r}
pivot_long_data = data %>%
  # Select and rename data
  select(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS,
    NIC
  ) %>% 
  # Create a numeric Chapter variable
  mutate(Chapter = substr(Chapter, 1, 2)) %>% 
  mutate(Chapter = as.integer(Chapter)) %>% 
  # Numeric filter
  filter(Chapter <= 10) %>% 
  # Character filter 
  filter(grepl("TYNESIDE", CCG)) %>% 
  filter(Practice != "UNIDENTIFIED DOCTORS") %>% 
  # Using ifelse() and mutate()
  mutate(Chapters = ifelse(Chapter <= 5, "1-5", "6-10")) %>% 
  # Group by new 'Chapters' column, to sum() items
  group_by(CCG, Practice, Chapters) %>% 
  summarise(
    Items = sum(Items),
    NIC = sum(NIC)
    ) %>% 
  ungroup() %>% 
  # Pivot 'long'
  pivot_longer(cols = c("Items", "NIC"), names_to = "Metric", values_to = "Value") %>% 
  # Format cell appearance
  mutate(CCG = stringr::str_to_title(CCG)) %>% 
  mutate(Practice = stringr::str_to_title(Practice))
```

## Dataset 2.7: Replace single string instance

If we wanted, we could then replace every instance of 'Ccg' for 'CCG'if we wanted, using gsub().

```{r}
pivot_long_data = data %>%
  # Select and rename data
  select(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS,
    NIC
  ) %>% 
  # Create a numeric Chapter variable
  mutate(Chapter = substr(Chapter, 1, 2)) %>% 
  mutate(Chapter = as.integer(Chapter)) %>% 
  # Numeric filter
  filter(Chapter <= 10) %>% 
  # Character filter 
  filter(grepl("TYNESIDE", CCG)) %>% 
  filter(Practice != "UNIDENTIFIED DOCTORS") %>% 
  # Using ifelse() and mutate()
  mutate(Chapters = ifelse(Chapter <= 5, "1-5", "6-10")) %>% 
  # Group by new 'Chapters' column, to sum() items
  group_by(CCG, Practice, Chapters) %>% 
  summarise(
    Items = sum(Items),
    NIC = sum(NIC)
    ) %>% 
  ungroup() %>% 
  # Pivot 'long'
  pivot_longer(cols = c("Items", "NIC"), names_to = "Metric", values_to = "Value") %>% 
  # Format cell appearance
  mutate(CCG = stringr::str_to_title(CCG)) %>% 
  mutate(Practice = stringr::str_to_title(Practice)) %>% 
  # Replace 'CCG' for 'Ccg'
  mutate(CCG = gsub("Ccg", "CCG", CCG))

gc()
```



## Dataset 3 -------------------------------------------------------------------

## Dataset 3.1: Rename and summarise data

First we will rename columns and then sum() items by CCG and Practice.

```{r}
case_when_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Items = ITEMS
  ) %>% 
  # Aggregate Data 
  group_by(CCG, Practice) %>% 
  summarise(Items = sum(Items)) %>% 
  ungroup()
```

## Dataset 3.2: Creating a new column using case_when()

Previously we used ifelse() and mutate() to create a new column.
An ifelse() creates 2 categories or outputs.
A case_when() can create as many categories as you want.
It is quite simple, even if the syntax appears odd at first.

```{r}
case_when_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Items = ITEMS
  ) %>% 
  # Aggregate Data 
  group_by(CCG, Practice) %>% 
  summarise(Items = sum(Items)) %>% 
  ungroup() %>% 
  # Mutate Aggregated Data
  mutate(
    `Item Volume` = case_when(
      Items < 10000 ~ "Low",
      Items >= 10000 & Items <= 20000 ~ "Medium ",
      Items > 20000 ~ "High"
    )
  )
```

## Dataset 3.3: Re-aggregate data by new column

Next, count the number of 'low volume', 'medium volume' and 'high volume' practices per CCG.
Note, this requires count(), rather than sum().
This is becuase:

- sum() is used with numerical values
- count() is used with categories (such as character values).

```{r}
case_when_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Items = ITEMS
  ) %>% 
  # Aggregate Data 
  group_by(CCG, Practice) %>% 
  summarise(Items = sum(Items)) %>% 
  ungroup() %>% 
  # Mutate Aggregated Data
  mutate(
    `Item Volume` = case_when(
      Items < 10000 ~ "Low",
      Items >= 10000 & Items <= 20000 ~ "Medium",
      Items > 20000 ~ "High"
    )
  ) %>% 
  # Aggregate Data for a Second time
  group_by(CCG) %>% 
  count(`Item Volume`, name = "Practice Count") %>% 
  ungroup() 
```

## Dataset 3.4: Order new columna as a factor

The Item Volume column is ordered alphabetically.
We want to specify the order differently.
First, we need to turn this into a 'factor' (categorical variable).
We can then specify the 'levels' of the factor.
If we use arrange(), it will then be presented as we want.

```{r}
case_when_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Items = ITEMS
  ) %>% 
  # Aggregate Data 
  group_by(CCG, Practice) %>% 
  summarise(Items = sum(Items)) %>% 
  ungroup() %>% 
  # Mutate Aggregated Data
  mutate(
    `Item Volume` = case_when(
      Items < 10000 ~ "Low",
      Items >= 10000 & Items <= 20000 ~ "Medium",
      Items > 20000 ~ "High"
    )
  ) %>% 
  # Aggregate Data for a Second time
  group_by(CCG) %>% 
  count(`Item Volume`, name = "Practice Count") %>% 
  ungroup() %>% 
  # Turn Items to factor, order factor levels and arrange data
  mutate(`Item Volume` = factor(`Item Volume`, levels = c("Low", "Medium", "High"))) %>% 
  arrange(CCG, `Item Volume`)

gc()
```

# Dataset 4 --------------------------------------------------------------------

## Dataset 4.1: Rename columns and aggregate data

We will sum() items, by CCG, Practice and BNF Chapter.
You don't need to select columns if you summarise() the data, as this determines which columns are retained.

```{r}
pivot_wide_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS
  ) %>% 
  # Aggregate Data 
  group_by(CCG, Practice, Chapter) %>% 
  summarise(Items = sum(Items)) %>% 
  ungroup()
```

## Dataset 4.2: Complete zero value entries

We might want to know which Practice-Chapter combinations have zero items.
We can use complete() to do this.
This makes explicit every Practice-Chapter combination.

We do *not* want every CCG-Practice-Chapter combination.
This is because a practice does not appear in every CCG (obviously).

```{r}
pivot_wide_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS
  ) %>% 
  # Aggregate Data 
  group_by(CCG, Practice, Chapter) %>% 
  summarise(Items = sum(Items)) %>% 
  ungroup() %>% 
  # Complete Practice-Chapter combinations
  complete(Practice, Chapter, fill = list(Items = 0))
```

## Dataset 4.3: Fill in CCG values

We only wanted to complete() Practice-Chapter combinations, and fill in the new values as zero.
We need to filll in the missing CCG values that resulted from this.
We can use fill() to do this.

If we group_by() Practice, this will fill in a missing CCG value, with the next one that appears for that practice. 

```{r}
pivot_wide_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS
  ) %>% 
  # Aggregate Data 
  group_by(CCG, Practice, Chapter) %>% 
  summarise(Items = sum(Items)) %>% 
  ungroup() %>% 
  # Complete Practice-Chapter combinations
  complete(Practice, Chapter, fill = list(Items = 0)) %>% 
  # Now fill in the missing CCG values (due to using complete)
  group_by(Practice) %>% 
  fill(CCG) %>% 
  ungroup()
```

## Dataset 4.4: Fill 'up' remaining CCG values

We can 'up', in addition to the default fill(), which goes 'down'.

```{r}
pivot_wide_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS
  ) %>% 
  # Aggregate Data 
  group_by(CCG, Practice, Chapter) %>% 
  summarise(Items = sum(Items)) %>% 
  ungroup() %>% 
  # Complete Practice-Chapter combinations
  complete(Practice, Chapter, fill = list(Items = 0, NIC = 0, Quantity = 0)) %>% 
  # Now fill in the missing CCG values (due to using complete)
  group_by(Practice) %>% 
  fill(CCG) %>% 
  # Remember to fill 'up' as well
  fill(CCG, .direction = "up") %>%  
  ungroup()
```

## Dataset 4.5: Pivot Wide

We can use pivot_wide, so that each chapter is now a new column.
The new columns will be populated from values in the 'Items' column.

```{r}
pivot_wide_data = data %>% 
  # Rename Columns
  rename(
    CCG = PCO_NAME,
    Practice = PRACTICE_NAME,
    Chapter = BNF_CHAPTER_PLUS_CODE,
    Items = ITEMS
  ) %>% 
  # Aggregate Data 
  group_by(CCG, Practice, Chapter) %>% 
  summarise(Items = sum(Items)) %>% 
  ungroup() %>% 
  # Complete Practice-Chapter combinations
  complete(Practice, Chapter, fill = list(Items = 0)) %>% 
  # Now fill in the missing CCG values (due to using complete)
  group_by(Practice) %>% 
  fill(CCG) %>% 
  # Remember to fill 'up' as well
  fill(CCG, .direction = "up") %>%  
  ungroup() %>% 
  # Arrange the data and pivot Wide 
  arrange(CCG, Practice, Chapter) %>% 
  pivot_wider(names_from = "Chapter", values_from = "Items")
```

## Collate all datasets into a spreadsheet and save

There are many R packages that let you read and write Excel files.
The `openxlsx` package allows you to save multiple dataframes as a single worksheet easily.
Other packages allow you to finely format a worksheet's appearance (such as bold text and cell shading etc).

First we must combine all of the dataframes into a 'list'.
A list is just a collection of objects, such as dataframes.
We can then save this list as an Excel file.
You must set a working directory, to tell R where you want the output to saved at.

```{r}
list_of_datasets = list(
  mutated_data,
  pivot_wide_data,
  case_when_data,
  pivot_long_data
)

setwd("C:/Users/adnsh/OneDrive - NHS Business Services Authority/Desktop/")
write.xlsx(list_of_datasets, file = "COFFEE_AND_CODING_EXCEL_TO_R_PART_ONE.xlsx")
```


## Summary

During this Coffee & Coding, we have learned about and used the following function:

Base R functions:

- substr()
- round()
- ifelse()
- gsub()
- grepl()
- as.integer()
- factor()
- paste0()

Dplyr functions:

- select()
- rename()
- filter()
- arrange()
- mutate()
- group_by()
- summarise()
- count()
- ungroup()
- case_when()

Tidyr functions:

- complete()
- fill()
- pivot_wider()
- pivot_longer()

Stringr functions:

- str_to_title()


## Next Steps

Using R instead of Excel for some tasks, will take a bit of time initially, but in the long run will save you a *lot* of time.
If you plan on doing, here are some recommended next steps:

1. First identify a repetitive task in Excel, for example that you on do on a weekly or monthly (or even daily) basis.

2. Write down *all* of the steps that you do (e.g. even renaming columns), that gets you from start to finish.

3. Then try convert each of steps into R code.

4. If you get stuck, look at this cheatsheet: https://www.rstudio.com/resources/cheatsheets/

5. If you are still stuck, post a question here: https://teams.microsoft.com/l/channel/19%3a64ee34434208492bad218f7bfe9cd47b%40thread.tacv2/Analytics%2520Queries?groupId=ccfef765-332b-4b0c-b2b7-dabc1d8e9148&tenantId=cf6d0482-86b1-4f88-8c0c-3b4de4cb402chttps://teams.microsoft.com/l/channel/19%3a64ee34434208492bad218f7bfe9cd47b%40thread.tacv2/Analytics%2520Queries?groupId=ccfef765-332b-4b0c-b2b7-dabc1d8e9148&tenantId=cf6d0482-86b1-4f88-8c0c-3b4de4cb402c

6. Get in touch at adnan.shroufi@nhs.net if you want something particular covered at the upcoming 'Excel -> R: Part Two' Coffee & Coding session.

Thanks!
