---
title: "Binary classification with R and Python"
author: "Kayoung Goffe"
date: "04/05/2022"
output: pdf_document
---

```{r setup, include=FALSE}
library(reticulate)
library(mlbench)
library(magrittr)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)

```

## Reticulate

The {reticulate} package integrates Python with R. 
We are going to run simple machine learning (binary classification) 


## Python configuration 

we need to create conda environment and install few packages. `conda_list()` function returns an R `data.frame`, with  
name giving the name of the associated environment, and Python giving the path to the Python binary associated with that environment.

```{r check conda list}

conda_list()

```
Why do we create our own conda environment? 

Sometimes you want to have very specific version of Python or related libraries for your project. From my limited experience working on MSc Machine Learning project, I have encountered the breaking changes error. (For example, Python version was not compatible with tensorflow library). The environment consists of a certain Python version and relevant packages.If you share your work with the environment, it will not cause breaking change error problem. 

Let's create our own conda environment using `conda_create(cc_env)` (it can)


```{r create my own conda environment}

# define our environment name
cc_env <- "coffee_coding_reticulate"

# create conda environment
conda_create(cc_env)

# check the conda environment
conda_list()
# We can see that new environment has been created.
```
Now we added new environment. We will use this environment and let's install python packages. 

```{r install python libraries to new environment}

# indicate that we want to use the environment we have just created
use_condaenv("coffee_coding_reticulate")

# install few most commonly used python packages; pandas, numpy, 
conda_install("coffee_coding_reticulate", "pandas")
conda_install("coffee_coding_reticulate", "numpy")
conda_install("coffee_coding_reticulate", "seaborn")
conda_install("coffee_coding_reticulate", "scikit-learn")

```
You can use `py_install()` function as well. It will require to install miniconda in your system. We are not using it today as we don't want to install while using binder. 

So far we have 1) created our own environment called 'coffee_coding_reticulate' and in that environment, we have installed four Python packages. Let's import python libraries in our R environment.

If you have anaconda installed, you will be able to see the new environment in Anaconda. You can install Python libraries through Anaconda if you want. I will show you on my local machine as I have installed anaconda. 


```{r read diabetes data from mlbench}

data(PimaIndiansDiabetes2)
head(PimaIndiansDiabetes2)

# Remove NA values and save as diabetes object
diabetes <- na.omit(PimaIndiansDiabetes2)

# splitting data into features and predicted variable

X <- diabetes[,1:8]
y <- data.frame(diabetes[,9]) %>% 
  dplyr::mutate(outcome = ifelse(`diabetes...9.`=='neg',0,1)) %>% 
  select(outcome)

```

Quick note on each variable:
* Pregnant: Number of times pregnant
* Glucose: Plasma glucose concentration (glucose tolerance test)
* Pressure: Diastolic blood pressure (mm Hg)
* Triceps: Skinfold thickness (mm)
* Insulin: 2-Hr serum insulin (mu U/ml)
* Mass: Body mass index (weight in Kg/ (height in m)² )
* Pedigree: Diabetes pedigree function
* Age: Age (years)

### Casting R object to Python using {r_to_py()}

You can try various ways of using R & Python together! I will show two different ways. First, we will pass R to Python object using `{r_to_py()}`. In this way, we can use Python libraries.


```{r import Python libraries}

# Python essential numpy, pandas are imported
numpy <- import('numpy')
pandas <- import('pandas')

# scikit-learn libraries
skl_preprocessing <- import("sklearn.preprocessing")
skl_model_selection <- import("sklearn.model_selection")
skl_linear_model <- import("sklearn.linear_model")
skl_metrics <- import("sklearn.metrics")

```










 

