---
title: "Binary classification with R and Python"
author: "Kayoung Goffe"
date: "04/05/2022"
output: pdf_document
---

```{r setup, include=FALSE}
library(reticulate)
knitr::opts_chunk$set(echo = TRUE)

```

## Reticulate

The {reticulate} package integrates Python with R. 
We are going to run simple machine learning (binary classification) 


## Python configuration 

we need to create conda environment and install few packages. `conda_list()` function returns an R `data.frame`, with  
name giving the name of the associated environment, and Python giving the path to the Python binary associated with that environment.

```{r check conda list}

conda_list()

```
Why do we create our own conda environment? 

Sometimes you want to have very specific version of Python or related libraries for your project. From my limited experience working on MSc Machine Learning project, I have encountered the breaking changes error. (For example, Python version was not compatible with tensorflow library). The environment consists of a certain Python version and relevant packages.If you share your work with the environment, it will not cause breaking change error problem. 

Let's create our own conda environment using `conda_create(cc_env)` (it can)


```{r create my own conda environment}

# define our environment name
cc_env <- "coffee_coding_reticulate"

# create conda environment
conda_create(cc_env)

# check the conda environment
conda_list()
```
Now we added new environment. We will use this environment and let's install python packages. 

```{r install python libraries}

# indicate that we want to use the environment we have just created
use_condaenv("coffee_coding_reticulate")

# install few most commonly used python packages; pandas, numpy, 
conda_install("coffee_coding_reticulate", "pandas")
conda_install("coffee_coding_reticulate", "numpy")
conda_install("coffee_coding_reticulate", "seaborn")
conda_install("coffee_coding_reticulate", "scikit-learn")

```
You can use `py_install()` function as well. It will require to install miniconda in your system. We are not using it today as we don't want to install while using binder. 

So far we have 1) created our own environment called 'coffee_coding_reticulate' and in that environment, we have installed four Python packages. Let's import python libraries in our R environment.

```{r using python in R importing libraries}


```










 

