---
title: "Flights from NYC - 2013"
author: "`r Sys.getenv('USERNAME')`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  html_document:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Delays report for carrier -  `r carrier_nm`

This is a report for all flights that departed NYC airports for the carrier - `r carrier_nm`.

## Delayed departures

```{r carrier_delays}

carrier_delays_df %>% 
  hchart(type = "line", hcaes(x = month_name, y = delayed_flights_per)) %>% 
  # we can make the title of the chart dynamic, to update automatically
  hc_title(text = paste0("Monthly percentage of delayed flights for carrier <span style='color: red;'><b>", carrier_nm, "</b></span>")) %>% 
  hc_xAxis(title = list(text = "Date"),
           type = "category") %>% 
  hc_yAxis(title = list(text = "Delayed flights (%)"),
           min = 0,
           max = 100)
```

## Average delay

```{r carrier_avg_delays}

carrier_avg_delays_df %>% 
  hchart(type = "line", hcaes(x = month_name, y = avg_delay)) %>% 
  # we can make the title of the chart dynamic, to update automatically
  hc_title(text = paste0("Average delay time <i>in minutes</i> for carrier <span style='color: red;'><b>", carrier_nm, "</b></span>")) %>% 
  hc_xAxis(title = list(text = "Date")) %>% 
  hc_yAxis(title = list(text = "Average delay time (mins)"),
           min = 0,
           max = 100)
```

## Most delayed flights

The flight with the longest delay for each month and most likely to have passengers claiming compensation are:

* January: __`r flights2 %>% filter(carrier == carrier, month == 1) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* February: __`r flights2 %>% filter(carrier == carrier, month == 2) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* March: __`r flights2 %>% filter(carrier == carrier, month == 3) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* April: __`r flights2 %>% filter(carrier == carrier, month == 4) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* May: __`r flights2 %>% filter(carrier == carrier, month == 5) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* June: __`r flights2 %>% filter(carrier == carrier, month == 6) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* July: __`r flights2 %>% filter(carrier == carrier, month == 7) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* August: __`r flights2 %>% filter(carrier == carrier, month == 8) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* September: __`r flights2 %>% filter(carrier == carrier, month == 9) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* October: __`r flights2 %>% filter(carrier == carrier, month == 10) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* November: __`r flights2 %>% filter(carrier == carrier, month == 11) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
* December: __`r flights2 %>% filter(carrier == carrier, month == 12) %>% filter(dep_delay == max(dep_delay, na.rm = TRUE)) %>% pull(tailnum)`__
