---
title: "An Introduction to Test-Driven Development (TDD)"
output: html_document
date: "2026-02-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  error = TRUE  # This allows the document to continue knitting even if a chunk errors
)

library(testthat)
```

The kata presented live in the session uses python to demonstrate the RED-GREEN-REFACTOR cycle. Here it is presented using R. Note that there is no general best order for implementing tests and features. It depends on the specification. Here we will proceed in the same order as the features are presented.

## The FizzBuzz kata

TODO: Use same definition as Ian will show live
TODO: Don't specify the language used

Fizz Buzz Problem involves that given an integer n, for every integer i <= n, the task is to write a program to print,

- 'Fizz' if i is divisible by 3,
- 'Buzz' if i is divisible by 5
- 'FizzBuzz' if i is divisible by 3 and 5,
- 'i' as a string, if none of the conditions are true.

TODO: Decide if we want just specify returning a list of strings 'up front', rather than printing?
Because testing a printed output would make it more complicated to show what we do in TDD, we will modify the specification to simply return a list of strings.

E.g. If you input '5', it should return `["1", "2", "Fizz", "3", "Buzz"]`

## Template test

When creating a new test, it is good to follow the 3 'A's - Assign, Act, Assert. A simple template for tests is below.

```r
should_output <- function() {
  # Assign
  
  
  # Act
  
  
  # Assert
  
}
```

## Simplest function

We can create the simplest possible function before starting the TDD process. That is just a function with no arguments and an empty body. This function can be called, but does nothing yet.

```{r}
fizzbuzzer <- function() {}

print(fizzbuzzer())
```

## Test 1 - "Fizz"

The first feature we will test is the "Fizz" part. We should expect a value of "Fizz" to be returned, when input is divisible by 3. Note that we are not implementing the entire specification here. We focus on one atomic feature. Once the basic features are implemented we will address the final output format (a list of strings).

```{r}
should_output_fizz_if_divisible_by_3 <- function() {
  # Assign
  input <- 3
  
  # Act
  output <- fizzbuzzer(input)
  
  # Assert
  expect_equal(output, "Fizz")
}
```

Running this results in an error. This is because our function has no argument, but in the 'Act' part of the test we call it with an argument.

```{r}
should_output_fizz_if_divisible_by_3()
```
This is a RED result, so we need to fix the function to get a successful GREEN result. With TDD, you should look to make only the minimal changes needed to get a pass. The idea is that a function is built up, in small well-tested increments, until the specification is met. The minimal change here is to add an argument for the number.
 
```{r}
fizzbuzzer <- function(number) {}
```

Running this again results in an error. This is because we expected the output to be "Fizz", but what we got was `NULL`. Another RED!

```{r}
should_output_fizz_if_divisible_by_3()
```

The simplest fix now is to simply output "Fizz" always. This might seem a waste of time! But consider the difference between writing 50 lines of code before running a test that fails, and writing 1 or a few lines before testing. In the latter case we can debug much faster as we know exactly where something went wrong.

```{r}
fizzbuzzer <- function(number) {
  "Fizz"
}
```

Now we get our first GREEN result. Most test libraries follow 'no news is good news', so we get nothing, telling us all is well.

```{r}
should_output_fizz_if_divisible_by_3()
```

You might be thinking we cheated to get that result, as we always return "Fizz". So we now need to REFACTOR our function to closer match the specification.

The modulus operator, `%%`, is the best way to test that one number is divisible by another. Given numbers `a` and `b`, it gives back the remainder of the division `a / b`, which is 0 if `a` is divisible by `b`.

```{r}
fizzbuzzer <- function(number) {
  if ((number %% 3) == 0) {
    return("Fizz")
  }
}
```

We get no 'news' again, so we have a GREEN result. This means we now have the behaviour for a number divisible by 3, as called for in the specification.

```{r}
should_output_fizz_if_divisible_by_3()
```

## Test 2 - "Buzz"

The second feature in the specification is *very* similar to what we just tested. We can replace the 'divisible by 3', with 'divisible by 5', and the output is "Buzz", rather then "Fizz". In such clearly similar cases, you could just copy a previous test and make the minor adjustments necessary. Especially when first learning TDD, it is recommended to still be methodical and build the test from scratch, using the 3 'A's.

```{r}
should_output_buzz_if_divisible_by_5 <- function() {
  # Assign
  input <- 5
  
  # Act
  output <- fizzbuzzer(input)
  
  # Assert
  expect_equal(output, "Buzz")
}
```

We get a RED result.

```{r}
should_output_buzz_if_divisible_by_5()
```

This is one we have seen before, and we can REFACTOR our function similarly to get the test to pass.

```{r}
fizzbuzzer <- function(number) {
  if ((number %% 3) == 0) {
    return("Fizz")
  }
  if ((number %% 5) == 0) {
    return("Buzz")
  }
}
```

And now we have a GREEN result. We now have 2 of the features covered.

```{r}
should_output_buzz_if_divisible_by_5()
```

## Test 3 - "FizzBuzz"

The next feature in the specification is a mishmash of the previous 2 we have tests for. Our input needs to be divisible by both 3 and 5 to test this, and we should expect "FizzBuzz" to be the result.

```{r}
should_output_fizzbuzz_if_divisible_by_3_and_5 <- function() {
  # Assign
  input <- 3 * 5
  
  # Act
  output <- fizzbuzzer(input)
  
  # Assert
  expect_equal(output, "FizzBuzz")
}
```

Handling this requires some care, because the conditions we already check in our function are also conditions involved in the "FizzBuzz" feature. Notably, the check for this combination must come *before* the individual checks.

```{r}
fizzbuzzer <- function(number) {
  if (((number %% 3) == 0) & ((number %% 5) == 0)) {
    return("FizzBuzz")
  }
  if ((number %% 3) == 0) {
    return("Fizz")
  }
  if ((number %% 5) == 0) {
    return("Buzz")
  }
}
```

This gets a GREEN result.

```{r}
should_output_fizzbuzz_if_divisible_by_3_and_5()
```

You might be thinking this could be improved, and it probably could be made more readable. One of the benefits of TDD is that once a test is in place, you have a very quick and easy way to check that behaviour (**not** implementation) is maintained when refactoring. Note that another benefit of TDD is that it leads to 'lean' code, as the developer writes minimal code to pass the tests. So always consider if any changes *do* improve your function. If in doubt, they are probably best not done!

So we could safely improve readability now, relying on the test to ensure we did not change the behaviour of the function.

```{r}
fizzbuzzer <- function(number) {
  is_divisible_by_3 <- (number %% 3) == 0
  is_divisible_by_5 <- (number %% 5) == 0
  
  if (is_divisible_by_3 & is_divisible_by_5) {
    return("FizzBuzz")
  }
  if (is_divisible_by_3) {
    return("Fizz")
  }
  if (is_divisible_by_5) {
    return("Buzz")
  }
}
```

After that refactoring of the conditions into clearly named variables, the function is more readable and it is very clear what is happening. The test stays GREEN.

```{r}
should_output_fizzbuzz_if_divisible_by_3_and_5()
```

## Test 4 - General case

When a number is *not* divisible by 3 or 5, we want to just output the number as a string.

```{r}
should_output_string_of_number_if_not_divisible_by_3_or_5 <- function() {
  # Assign
  input <- 2
  
  # Act
  output <- fizzbuzzer(input)
  
  # Assert
  expect_equal(output, "2")
}
```

We get a familiar RED result here.

```{r}
should_output_string_of_number_if_not_divisible_by_3_or_5()
```

As we have seen, the NULL is returned because we have not handled the case in our function yet. This can be considered a default case, so we just return the number in string format. This way, when all the previous conditions do not pass, we 'fall through' to the default case.

```{r}
fizzbuzzer <- function(number) {
  is_divisible_by_3 <- (number %% 3) == 0
  is_divisible_by_5 <- (number %% 5) == 0
  
  if (is_divisible_by_3 & is_divisible_by_5) {
    return("FizzBuzz")
  }
  if (is_divisible_by_3) {
    return("Fizz")
  }
  if (is_divisible_by_5) {
    return("Buzz")
  }
  
  as.character(number)
}
```

And we have a GREEN result.

```{r}
should_output_string_of_number_if_not_divisible_by_3_or_5()
```

## Test 5 - List output

Up to now, we have focused on handling of individual numbers. But the specification calls for a list of strings to be returned. This is an example of the 'divide and conquer' strategy (useful outside of coding too!). We broke the specifications down into small parts of the final set of features. Now it is time to bring everything together.

One way to go here would be to refactor our existing tests and `fizzbuzzer` to output a list. But this will mean essentially 'throwing away' our good and reliable tests.

TODO: Check with Ian that below is correct use of SR/SoC?

A better way is to use the principles of Single Responsibility and Separation of Concerns. This means having small functions, that work independently and do just one thing. This can result in more maintainable and modular code. Here, it also allows us to keep our existing tests, so our final test suite is more robust. 

The first thing we should do is rename `fizzbuzzer`. This function now represents the logic, without caring about how it is actually used. We need to refactor the existing tests also, and do this first.

```{r}
should_output_fizz_if_divisible_by_3 <- function() {
  # Assign
  input <- 3
  
  # Act
  output <- fizzbuzz(input)
  
  # Assert
  expect_equal(output, "Fizz")
}
```

```{r}
should_output_fizz_if_divisible_by_5 <- function() {
  # Assign
  input <- 5
  
  # Act
  output <- fizzbuzz(input)
  
  # Assert
  expect_equal(output, "Buzz")
}
```

```{r}
should_output_fizzbuzz_if_divisible_by_3_and_5 <- function() {
  # Assign
  input <- 3 * 5
  
  # Act
  output <- fizzbuzz(input)
  
  # Assert
  expect_equal(output, "FizzBuzz")
}
```

```{r}
should_output_string_of_number_if_not_divisible_by_3_or_5 <- function() {
  # Assign
  input <- 2
  
  # Act
  output <- fizzbuzz(input)
  
  # Assert
  expect_equal(output, "2")
}
```

```{r}
fizzbuzz <- function(number) {
  is_divisible_by_3 <- (number %% 3) == 0
  is_divisible_by_5 <- (number %% 5) == 0
  
  if (is_divisible_by_3 & is_divisible_by_5) {
    return("FizzBuzz")
  }
  if (is_divisible_by_3) {
    return("Fizz")
  }
  if (is_divisible_by_5) {
    return("Buzz")
  }
  
  as.character(number)
}
```

All that was changed was to rename `fizzbuzzer` to `fizzbuzz`. We still get GREEN tests.

```{r}
should_output_fizz_if_divisible_by_3()
should_output_fizz_if_divisible_by_5()
should_output_fizzbuzz_if_divisible_by_3_and_5()
should_output_string_of_number_if_not_divisible_by_3_or_5()
```

Now we need to write a test for our actual `fizzbuzzer` function, which will output a list of strings, as per specification.

```{r}
should_output_correct_list_for_input_of_15 <- function() {
  # Assign
  input <- 15
  
  # Act
  output <- fizzbuzzer(input)
  
  # Assert
  expect_equal(
    output,
    list("1", "2", "Fizz", "4", "Buzz", "Fizz", "7", "8", "Fizz", "Buzz", "11", "Fizz", "13", "14", "FizzBuzz")
  )
}
```

And we can now write the function that will complete implementation of the specification.

```{r}
fizzbuzzer <- function(number) {
  out <- list()
  
  for (i in 1:number) {
    out <- c(out, fizzbuzz(i))
  }
  
  out
}
```

Green result!

```{r}
should_output_correct_list_for_input_of_15()
```

Now we have a fully working and well-tested implementation it is a good time to consider if we want to REFACTOR. The `fizzbuzzer` function has some non-ideal issues. The main one is potentially very poor memory efficiency (the pattern used will result in the list being copied into a new memory location every iteration, which could get out of hand for large numbers). There is also a bug should 0 be input (`1:number` would become `c(1, 0)`). The idiomatic way of doing this in R would be using `lapply` for the list building, and `seq_len` for the sequence of numbers to process.

```{r}
fizzbuzzer <- function(number) lapply(seq_len(number), fizzbuzz)
```

Always run your tests after a refactor. Still GREEN.

```{r}
should_output_correct_list_for_input_of_15()
```

And just to check the output visually...

```{r}
cat(as.character(fizzbuzzer(15)), sep = ", ")
```