---
title: "An Introduction to Test-Driven Development (TDD)"
output: html_document
date: "2026-02-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(testthat)
```

## The FizzBuzz kata

Fizz Buzz Problem involves that given an integer n, for every integer i <= n, the task is to write a Python program to print,

- 'Fizz' if i is divisible by 3,
- 'Buzz' if i is divisible by 5
- 'FizzBuzz' if i is divisible by 3 and 5,
- 'i' as a string, if none of the conditions are true.

Because testing a printed output would make it more complicated to show what we do in TDD, we will modify the function to create a list of elements to print. E.g.

If you input '5', it should return `["1", "2", "Fizz", "3", "Buzz"]`, while printing

```r
1
2
"Fizz"
4
"Buzz"
```

## Template test

```{r}
should_output <- function() {
  # Assign
  
  
  # Act
  
  
  # Assert
  
}
```

## Test 1 - "Fizz"

```{r}
fizzbuzzer <- function() {} # Start - empty function
```

```{r}
should_output_fizz_if_divisible_by_3 <- function() {   # NOTE: naming considerations, parameter in name?
  # Assign
  input <- 3
  
  # Act
  output <- fizzbuzzer(input)
  
  # Assert
  expect_equal(output, "Fizz")
}
```

```{r}
should_output_fizz_if_divisible_by_3() # RED: no argument in functon
```

```{r}
fizzbuzzer <- function(number) {} # REFACTOR: add argument
```

```{r}
should_output_fizz_if_divisible_by_3() # RED: output not as expected
```

```{r}
fizzbuzzer <- function(number) { # REFACTOR: change output to what test expects
  return ("Fizz")
}
```

```{r}
should_output_fizz_if_divisible_by_3() # GREEN - but note the behaviour being tested is not yet as specified
```

```{r}
fizzbuzzer <- function(number) { # REFACTOR: change behaviour of function to match specification
  # if (number %% 3) { # Imagine making this mistake...
  if ((number %% 3) == 0) {
    return("Fizz")
  }
  
  as.character(number)
}
```

```{r}
should_output_fizz_if_divisible_by_3() # COMPLETED CYCLE
```

## Test 2 -  default "i"

We have 2 branches in our `fizzbuzzer` function, but the first test only tests one of them. Where possible, testing **every** branch is better.

```{r}
should_output_input_as_string <- function() {
  # Assign
  input <- 2
  
  # Act
  output <- fizzbuzzer(input)
  
  # Assert
  assert_equal(output, as.character(input)) # No 'magic' number, reliably uses input to create what we expect.
                                            # BUT - could result in hidden bug? Does DRY apply to test assertions?
  
  assert_equal(output, "2")                 # Explicit expectation. If input changes, this does not, but we will 
                                            # quickly see due to failing test. Safer?
}
```
