---
title: "Introduction to Regular Expressions"
author: "Mihail Garbuzov"
date: "2023-06-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
library(dplyr)
library(stringr)
```

## Exercise problem solutions

# People's names

Write a pattern that matches a person's forename and surname as part of a wider text. What disadvantages would this method have?

```{r}
"In 1831, Charles Darwin embarked on a voyage aboard a ship employed as a naturalist." |>
  str_view("[:upper:][:lower:]+\\s[:upper:][:lower:]+")

# Won't match a full name if any middle names are present
# Will unintentionally match non-person's names, e.g. Crystal Palace
```

# Assignment numbers and NINs

In ESR data, employees have assignment numbers with role IDs, which have a pattern `11111111-11` (where the role ID after the hyphen is optional) and national insurance numbers, which have a pattern `AA111111A`. In a pension opt-out dataset we have had to work with in the past, there was an `employee_id` column, which included one or the other of these. Write patterns that label the type of number in `employee_id`.

```{r}
df <- tibble(
  employee_id = c("13654789-01", "16345628", "ZX364578A", "07599999999")
)

df |> mutate(
  employee_id_type = case_when(
    str_detect(employee_id, "^\\d{8}(-\\d{2})?$") ~ "Assignment No.",
    str_detect(employee_id, "^\\w{2}\\d{6}\\w{1}$") ~ "NIN",
    T ~ "Other"
  )
)

```

# NHS numbers

Write a pattern that matches valid NHS numbers only. A valid number consists of 10 digits, but there are some invalid dummy numbers, such as 0000000000, 1111111111, etc.

```{r}

df <- tibble(
  nhs_no = c("7893544451", "0000000000")
)

df |> mutate(
  nhs_no_type = case_when(
    str_detect(nhs_no, "^\\d{10}$") & !str_detect(nhs_no, "^(\\d)(\\1){9}$") ~ "Valid",
    T ~ "Invalid"
  )
)

```

# Email logins

Write a piece of code that extracts login names from email addresses that occur in a wider text. There are at least two methods of doing this.

```{r }
"My email is mihail.garbuzov@nhsbsa.nhs.uk." |>
  str_extract("([\\w\\.]+)@[\\w\\.]+\\w+", group = 1) # Using a capturing group

"My email is mihail.garbuzov@nhsbsa.nhs.uk." |>
  str_extract("[\\w\\.]+(?=@[\\w\\.]+\\w+)") # Using lookahead

```

# UK postcodes

UK postcodes may come in various formats, listed below

A0 0AA, A00 0AA, AA0 0AA, AA00 0AA, A0A 0AA, AA0A 0AA

Write a piece of code with one or more Regex patterns to valid postcodes. What is a limitation of this method?

```{r}
postcodes <- c(
  "A0 0AA",
  "A00 0AA",
  "AA0 0AA",
  "AA00 0AA",
  "A0A 0AA",
  "AA0A 0AA",
  # invalid formats included for testing
  "A0AA 0AA",
  "A00A 0AA"
)

str_view(postcodes, "[A-Z]{1,2}[0-9][A-Z0-9]? [0-9][A-Z]{2}", match = NA)

# Explanation: the first half begins with 1-2 letters followed by a digit,
# after which there is either a second digit, a letter or nothing.
# Two halves are separated by a space, which is optional, making the pattern more robust
# The second half is always strictly made of one digit and two letters

# Limitation is that while Regex may validate a format, it doesn't guarantee a 
# postcode is real and existing.
```
