---
title: "Highcharts - Introduction"
author: "Adnan Shroufi"
date: "23/11/2021"
output: html_document
---

### Introduction ---------------------------------------------------------------

## Introduction to Highcharter

This is an introduction to highcharter.
We will look at:

- Using the hchart() function
- Editing the appearance of a highchart
- Making a highchart interactive
- Using dplyr for data manipulation

We'll be using a small extract of data from the ODP to do this.

## Load Packages

Before we bring in the data downloaded from data ODP, we need to make sure have all the required packages installed.

```{r, message=F}
library(highcharter)
library(dplyr)
library(tidyr)
library(skimr)
```

## Load some Data downloaded from the ODP

To save time, data has been already extacted from the ODP.
We will load this directly, rather than 'live' through the API.
Below are the key features of the data:

- 12 months of data (2020)
- BNF Chapter 23 on Stoma Appliances (e.g. colostomy bags)
- Only for postcodes beginning with 'NE42' (the Prudhoe area)

```{r pressure, echo=FALSE}
data = readRDS("highcharter_intro.Rds")
```

### Data Overview --------------------------------------------------------------

## Summary Functions

We can have a quick look at the data and see that it's fairly straight forward, with just 6 columns.

```{r}
head(data)
skim(data)
```

## Further Inspection of Category-based Variables

We can inspect some of the category-based variables.
We can see that most of records relate to prescribing from the Prudhoe Medical Group.

```{r}
table(data$practice_name)
```

Skin fillers and protectives, adhesive removers and colostomy bags were the most prescribed stoma appliances.

```{r}
table(data$chemical_substance_bnf_descr)
```

If we pitch the above 2 variables against each other, we can get a handy 2-way summary table.

```{r}
table(data$chemical_substance_bnf_descr, data$practice_name)
```

### Creating Charts using Highcharter ------------------------------------------

## Introduction to hchart()

In highcharter there are a couple of ways of creating plots, either using hchart() or with individual series.
This introduction will focus on hchart, as it is the easiest way to create nice looking plots with highcharter.

The hchart() function is very similar to the ggplot() function, although there are a couple of differences.
In ggplot you specify the aes() and then on a following line specify the plot type.

E.g. ggplot(data, aes(x = practice_name, y = items))+
      geom_col()

Whereas with hchart you specify the plot type within the initial function.

E.g. hchart(data, type = "column", hcaes(x = practice_name, y = items))

The hchart function works in a similar way to ggplot, with the hcaes() functioning similar to ggplot's aes().
Aside from that, while you '+' ggplot commands together, in highcharter you need to '%>%' them.
All in all, the way you chain multiple lines of code together is similar, in order to build plots of building complexity.
We'll be doing that in this introduction, starting with simple plots and then building them up.

We will be looking at 3 plots.
First up is a bar chart.

## Plot 1.1: Column/Bar Chart

The first plot is looked at all items across all 3 practices, grouped by the 12 'chemical substances' present in the data.
After we've grouped the data, the 12 chemical substances are put on the x-axis and the number of items on the y-axis.


```{r, message=F}
filtered_data = data %>% 
  group_by(chemical_substance_bnf_descr) %>% 
  summarise(items = sum(items)) %>% 
  ungroup()

filtered_data

hchart(filtered_data, type = "column", hcaes(x = chemical_substance_bnf_descr, y = items))
```

# Plot 1.2: 'Piping' Data straight into hchart

An alternative way of doing the above is to actually the aggregation and plot together.
You do can this by piping the aggregation code straight into hchart().
As a result, you don't need to insert any data object into the function, as the data is already being piped through.
We'll be doing this with hchart() for the rest of the introduction.
The below code returns the same as above, but as a single code-block.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr) %>% 
  summarise(items = sum(items)) %>% 
  ungroup() %>% 
  hchart(type = "column", hcaes(x = chemical_substance_bnf_descr, y = items))
```

## Plot 1.3: Arranging the Data

Before evven tweaking any highcharter code, we can order the bars.
This is using the arrange(...) function within dplyr.
If you used arrange(desc(...)), this would order the data from largest from smallest.
The would mean the tallest bar would be on the far left.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr) %>% 
  summarise(items = sum(items)) %>% 
  ungroup() %>% 
  arrange(items) %>% 
  hchart(type = "column", hcaes(x = chemical_substance_bnf_descr, y = items))
```

## Plot 1.4: Group by Practice

Putting the practice_name within the group_by allows us to see the number of items by chem-sub per practice.
Within the hcaes(), if we put in group = practice_name, this then separates each practice as a different group.
Within a "column" type plot, this defaults to a side-by-side plot.
This could be turned into a stacked bar without much work if needed.
That code is hashed-out below, as we haven't covered hc_plotOptions yet.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(items = sum(items)) %>% 
  ungroup() %>% 
  arrange(items) %>% 
  hchart(type = "column", hcaes(x = chemical_substance_bnf_descr, y = items, group = practice_name)) 
```

## Plot 1.5: Add a Tooltip

One of the (many) good things about highcharter, is the ease at which a tooltip can be implemented/formatted.
You can do this manually, as we'll do later.
Or if you just want each series (i.e. group = practice_name) to appear within the same tooltip, using hc_tooltip(table = T).

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(items = sum(items)) %>% 
  ungroup() %>% 
  arrange(items) %>% 
  hchart(type = "column", hcaes(x = chemical_substance_bnf_descr, y = items, group = practice_name)) %>% 
  hc_tooltip(table = T)
```

## Plot 1.6 Commplete the Data within the Chart

However, the tooltip appear incomplete.
This is because we need to generate zero values when a practice does not have any chem-sub types.
To generate these zero values, we can use complete().
The tooltip will now show these 'missing' values.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(items = sum(items)) %>% 
  ungroup() %>% 
  complete(chemical_substance_bnf_descr, practice_name, fill = list(items = 0)) %>% 
  arrange(items) %>% 
  hchart(type = "column", hcaes(x = chemical_substance_bnf_descr, y = items, group = practice_name)) %>% 
  hc_tooltip(table = T)
```

# Plot 1.7: Bar vs Column

Using the exact same code, we can change the plot to a horizontal bar chart, just by using type = "bar", rather than "column".

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(items = sum(items)) %>% 
  ungroup() %>% 
  complete(chemical_substance_bnf_descr, practice_name, fill = list(items = 0)) %>% 
  arrange(items) %>% 
  hchart(type = "bar", hcaes(x = chemical_substance_bnf_descr, y = items, group = practice_name)) %>% 
  hc_tooltip(table = T)
```

## Plot 2.1: Basic Line Chart

If type is changed to "line", this creates a line chart.
Obviously we neeed to use year_month for the x-axis here.
There is only 1 line, as the data is not grouped by practice.

```{r, message=F}
data %>% 
  group_by(year_month) %>% 
  summarise(nic = round(sum(nic))) %>% 
  ungroup() %>% 
  hchart(type = "line", hcaes(x = year_month, y = nic))
```

## Plot 2.2: Grouped Line Chart

If we add practice_name to the group_by and the hcaes(), we will create 1 line per practice.

```{r, message=F}
data %>% 
  group_by(year_month, practice_name) %>% 
  summarise(nic = round(sum(nic))) %>% 
  ungroup() %>% 
  hchart(type = "line", hcaes(x = year_month, y = nic, group = practice_name))
```

# Plot 2.3: Add Tooltip and Complete Data

As with the previous plot, we can generate the zero values using complete and add a tooltip-table.

```{r, message=F}
data %>% 
  group_by(year_month, practice_name) %>% 
  summarise(nic = round(sum(nic))) %>% 
  ungroup() %>% 
  complete(year_month, practice_name, fill = list(nic = 0)) %>% 
  hchart(type = "line", hcaes(x = year_month, y = nic, group = practice_name)) %>% 
  hc_tooltip(table = T)
```

## Plot 2.4: Add a Zoom

Another quick win.
Using hc_chart(zoomType = "..."), we can add a zoom.
"xy" means the zoom box can be dragged both horizontally and vertically.

```{r, message=F}
data %>% 
  group_by(year_month, practice_name) %>% 
  summarise(nic = round(sum(nic))) %>% 
  ungroup() %>% 
  complete(year_month, practice_name, fill = list(nic = 0)) %>% 
  hchart(type = "spline", hcaes(x = year_month, y = nic, group = practice_name)) %>% 
  hc_tooltip(table = T) %>% 
  hc_chart(zoomType = "xy")
```

## Plot 2.5: Edit Zoom Type and Manually Specify Tooltip

With a line chart, you may want a zoom that only moves horizontally.
This means a zoom on a particular year_month will not exclude any data.
This can be done easilt through zoomType = "x", rather than "xy".

To manually specify the tooltip we use 'pointFormat = ' to format the tooltip appearance.
We can use HTML within this to fine tune the appearance.

'<b>' means begin bold text.
'</b>' means end bold text.
This means the word 'NIC:' will be in bold below.

In contrast to the hard-coded text above, we can choose appears dynamically, as the cursor moves.
{point.vars_name} determines this.

{point.items} = items
{point.year_month} = year_month
And so on.

```{r, message = F}
data %>% 
  group_by(year_month, practice_name) %>% 
  summarise(nic = round(sum(nic))) %>% 
  ungroup() %>% 
  complete(year_month, practice_name, fill = list(nic = 0)) %>% 
  hchart(type = "spline", hcaes(x = year_month, y = nic, group = practice_name)) %>% 
  hc_tooltip(
    pointFormat = "<b>NIC:</b> {point.nic}"
    )
```

## Plot 2.6: Continue to Develop Chart Tooltip

We can now add further items to the tooltip.
'<br>' = line break.
This means we can spread a tooltip over multiple lines.

```{r, message = F}
data %>% 
  group_by(year_month, practice_name) %>% 
  summarise(nic = round(sum(nic))) %>% 
  ungroup() %>% 
  complete(year_month, practice_name, fill = list(nic = 0)) %>% 
  hchart(type = "spline", hcaes(x = year_month, y = nic, group = practice_name)) %>% 
  hc_chart(zoomType = "x") %>% 
  hc_tooltip(
    pointFormat = "<b>Year Month:</b> {point.year_month}<br>
    <b>Practice Name:</b> {point.practice_name}<br>
    <b>NIC:</b> {point.nic}"
  )
```

## Plot 2.7: Mutate Additional Variables Specifically for Tooltip

The mutate() function either creates, or overrides, a variable.
Using substr(), inside a mutate, we can then take a substring of year_month to get the year and month alone. 
The tooltip code needs to be edited to reflect this.

```{r, message = F}
data %>% 
  group_by(year_month, practice_name) %>% 
  summarise(nic = round(sum(nic))) %>% 
  ungroup() %>% 
  complete(year_month, practice_name, fill = list(nic = 0)) %>% 
  mutate(
    year = substr(year_month, 1, 4),
    month = substr(year_month, 5, 6)
  ) %>% 
  hchart(type = "spline", hcaes(x = year_month, y = nic, group = practice_name)) %>% 
  hc_chart(zoomType = "x") %>% 
  hc_tooltip(
    pointFormat = "<b>Year:</b> {point.year}<br>
    <b>Month:</b> {point.month}<br>
    <b>Practice Name:</b> {point.practice_name}<br>
    <b>NIC:</b> {point.nic}"
  )
```

## Plot 2.8: Generate Month as words and Add £ to NIC

We can convert a month from a number to a word using month.name[...].
However, month,name[...] expects an integer, but the output of substr() is a character.
Therefore, we need to convert the substr month to an integer first.
We can also insert a '£' symbol into the tooltip.

```{r, message = F}
data %>% 
  group_by(year_month, practice_name) %>% 
  summarise(nic = round(sum(nic))) %>% 
  ungroup() %>% 
  complete(year_month, practice_name, fill = list(nic = 0)) %>% 
  mutate(
    year = substr(year_month, 1, 4),
    month = substr(year_month, 5, 6),
    month = month.name[as.integer(month)]
    ) %>% 
  hchart(type = "spline", hcaes(x = year_month, y = nic, group = practice_name)) %>% 
  hc_chart(zoomType = "x") %>% 
  hc_tooltip(
    pointFormat = "<b>Year:</b> {point.year}<br>
    <b>Month:</b> {point.month}<br>
    <b>Practice Name:</b> {point.practice_name}<br>
    <b>NIC:</b> £{point.nic}"
  )
```

## Plot 2.9: Finally, remove the Tooltip Header, add a Thousand Separator and change x-axis 

Within the manually specified tooltip, we need to remove the header.
This can be done using 'headerFormat = ""'.
A little bit of code can also generate a thousand separator.
We can also set month on the x-axis instead of year_month.

```{r, message=F}
data %>% 
  group_by(year_month, practice_name) %>% 
  summarise(nic = round(sum(nic))) %>% 
  ungroup() %>% 
  complete(year_month, practice_name, fill = list(nic = 0)) %>% 
  mutate(
    year_month = as.integer(year_month),
    year = substr(year_month, 1, 4),
    month = as.integer(substr(year_month, 5, 6)),
    month = month.name[month]
    ) %>% 
  hchart(type = "spline", hcaes(x = month, y = nic, group = practice_name)) %>% 
  hc_chart(zoomType = "x") %>% 
  hc_tooltip(
    headerFormat = "",
    pointFormat = "<b>Year:</b> {point.year}<br>
    <b>Month:</b> {point.month}<br>
    <b>Practice Name:</b> {point.practice_name}<br>
    <b>NIC:</b> £{point.nic:,2f}"
  )
```

## Plot 3.1: Basic Scatter Graph

'type = "scatter"' creates a basic scatter plot.
'group = ' determines the series within the data.

3 groups create 3 series.
Each series can be selected/de-selectedd from the legend.
Each series is coloured differently.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, group = practice_name))
```

## 3.2: Add a Zoom

Add a zoom, zooms are great for exploring data within a scatter plot.
Highcharter scatter plots with zoom, can be a great way to conduct EDA.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, group = practice_name)) %>% 
  hc_chart(zoomType = "xy") 
```

## Plot 3.3: Edit Tooltip & Add Z-Value

Although the tooltip highlights an entire series, the data displayed just refers to a single point.
You will need to manually create a tooltip for a scatter plot.

If we add 'z = ' to the plot, we can determine point size in relation to a variable within the data.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, z = total_quantity, group = practice_name)) %>% 
  hc_chart(zoomType = "xy") %>% 
  hc_tooltip(
    pointFormat = "<b>Chemical Substance:</b> {point.chemical_substance_bnf_descr}"
  )
```

## Plot 3.4: Further Edit Tooltip & Z-Value

We can build the tooltip up as like with a previous plot.

If we square the z-value, this will accentuate the difference in size in point size.
Squaring the values means the relative difference between the smallest and biggest values will be bigger. 

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, z = total_quantity^2, group = practice_name)) %>% 
  hc_chart(zoomType = "xy") %>% 
  hc_tooltip(
    pointFormat = "<b>Chemical Substance:</b> {point.chemical_substance_bnf_descr} <br>
    <b>NIC:</b> {point.nic} <br>
    <b>Items:</b> {point.items} <br>
    <b>Total Quantity:</b> {point.total_quantity}"
  )
```

## Plot 3.5: Change Series Colors

There are 3 series.
If we pass 3 colours to hc_colors, we can change each respective colour.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, z = total_quantity^2, group = practice_name)) %>% 
  hc_chart(zoomType = "xy") %>% 
  hc_tooltip(
    pointFormat = "<b>Chemical Substance:</b> {point.chemical_substance_bnf_descr} <br>
    <b>NIC:</b> £{point.nic:,.0f} <br>
    <b>Items:</b> {point.items:,.0f} <br>
    <b>Total Quantity:</b> {point.total_quantity:,.0f} "
  ) %>% 
  hc_colors(colors = c("darkblue", "lightblue", "purple"))
```

## Plot 3.6: Credits & Title

Credits need to be added to all highcharts plots, which is easily done.
A title can also be easily added.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, z = total_quantity^2, group = practice_name)) %>% 
  hc_chart(zoomType = "xy") %>% 
  hc_tooltip(
    pointFormat = "<b>Chemical Substance:</b> {point.chemical_substance_bnf_descr} <br>
    <b>NIC:</b> £{point.nic:,.0f} <br>
    <b>Items:</b> {point.items:,.0f} <br>
    <b>Total Quantity:</b> {point.total_quantity:,.0f}"
  ) %>% 
  hc_colors(colors = c("darkblue", "lightblue", "purple")) %>% 
  hc_credits(enabled = T) %>% 
  hc_title(text = "Total NIC and Items for GP Practices in Prudhoe in 2020")
```

## Plot 3.7: Fine Tune Title Appearance 

We can also use HTML within hc_title.
We can also change the font and font size easily.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, z = total_quantity^2, group = practice_name)) %>% 
  hc_chart(zoomType = "xy") %>% 
  hc_tooltip(
    pointFormat = "<b>Chemical Substance:</b> {point.chemical_substance_bnf_descr} <br>
    <b>NIC:</b> £{point.nic:,.0f} <br>
    <b>Items:</b> {point.items:,.0f} <br>
    <b>Total Quantity:</b> {point.total_quantity:,.0f}"
  ) %>% 
  hc_colors(colors = c("darkblue", "lightblue", "purple")) %>% 
  hc_credits(enabled = T) %>% 
  hc_title(
    text = "<b>Total NIC and Items for GP Practices in Prudhoe in 2020</b>",
    style = list(
      fontFamily = "arial",
      fontSize = 25
      )
  )
```

## Plot 3.8: Change Theme and Add Subtitle

We can add a subtitle too easily.
Again, HTML can be used, for example to create a line break.

You can also change the theme of a plot, which changes how its axes and background appears.
Note, some themes also change the font.
Custom themes can also be created, although we won't do that today.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, z = total_quantity^2, group = practice_name)) %>% 
  hc_chart(zoomType = "xy") %>% 
  hc_tooltip(
    pointFormat = "<b>Chemical Substance:</b> {point.chemical_substance_bnf_descr} <br>
    <b>NIC:</b> £{point.nic:,.0f} <br>
    <b>Items:</b> {point.items:,.0f} <br>
    <b>Total Quantity:</b> {point.total_quantity:,.0f}"
  ) %>% 
  hc_colors(colors = c("darkblue", "lightblue", "purple")) %>% 
  hc_credits(enabled = T) %>% 
  hc_title(
    text = "<b>Total NIC and Items for GP Practices in Prudhoe in 2020</b>",
    style = list(
      fontFamily = "arial",
      fontSize = 25
    )
  ) %>% 
  hc_subtitle(text = "Only Stoma Appliances Included <br> BNF Chapter 23") %>% 
  hc_add_theme(hc_theme_elementary())
```

## Plot 3.9: Add a Caption

We can also easily add a caption below a plot.
Again, this can contain HTML.
<em> and </em> allow us to italicise text.
We can also use the 'align=' and 'margin=' parameters to change the caption position, and the distance between the caption and the plot.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, z = total_quantity^2, group = practice_name)) %>% 
  hc_chart(zoomType = "xy") %>% 
  hc_tooltip(
    pointFormat = "<b>Chemical Substance:</b> {point.chemical_substance_bnf_descr} <br>
    <b>NIC:</b> £{point.nic:,.0f} <br>
    <b>Items:</b> {point.items:,.0f} <br>
    <b>Total Quantity:</b> {point.total_quantity:,.0f}"
  ) %>% 
  hc_colors(colors = c("darkblue", "lightblue", "purple")) %>% 
  hc_credits(enabled = T) %>% 
  hc_title(
    text = "<b>Total NIC and Items for GP Practices in Prudhoe in 2020</b>",
    style = list(
      fontFamily = "arial",
      fontSize = 25
    )
  ) %>% 
  hc_subtitle(text = "Only Stoma Appliances Included <br> BNF Chapter 23") %>% 
  hc_add_theme(hc_theme_elementary()) %>% 
  hc_caption(
    text = '<b>Size of points related to Total Quantity</b> <br><em>Use Tooltip to Check Total Quantity</em>',
    align = "center",
    margin = 20
    )
```

## Plot 3.10: Edit Axes

Axis text can be added easily enough.
Axis limits can be defined too within the same functions.

```{r, message=F}
data %>% 
  group_by(chemical_substance_bnf_descr, practice_name) %>% 
  summarise(
    nic = sum(nic),
    total_quantity = sum(total_quantity),
    items = sum(items)
  ) %>% 
  ungroup() %>% 
  hchart(type = "scatter", hcaes(x = nic, y = items, z = total_quantity^2, group = practice_name)) %>% 
  hc_chart(zoomType = "xy") %>% 
  hc_tooltip(
    pointFormat = "<b>Chemical Substance:</b> {point.chemical_substance_bnf_descr} <br>
    <b>NIC:</b> £{point.nic:,.0f} <br>
    <b>Items:</b> {point.items:,.0f} <br>
    <b>Total Quantity:</b> {point.total_quantity:,.0f}"
  ) %>%  
  hc_colors(colors = c("darkblue", "lightblue", "purple")) %>% 
  hc_credits(enabled = T) %>% 
  hc_title(
    text = "<b>Total NIC and Items for GP Practices in Prudhoe in 2020</b>",
    style = list(
      fontFamily = "arial",
      fontSize = 25
    )
  ) %>% 
  hc_subtitle(text = "Only Stoma Appliances Included <br> BNF Chapter 23") %>% 
  hc_caption(
    text = '<b>Size of points related to Total Quantity</b> <br><em>Use Tooltip to Check Total Quantity</em>',
    align = "center",
    margin = 20
  ) %>% 
  hc_add_theme(hc_theme_elementary()) %>% 
  hc_xAxis(
    min = 0,
    max = 7000,
    title = list(text = "Total Ingredient Cost (NIC) in £")
  ) %>% 
  hc_yAxis(
    min = 0,
    max = 200,
    title = list(text = "Total Items")
  )
```

### Review of 3 Charts

If we look at the final code for the 3 plot charts, while it looks fairly lengthy, if you break it down it's fairly simple.
Using a small number of functions we can build plots of increasing complexity.
The final thing, is to now review all of the code/functions used today (see powerpoint).

## End of Highcharter introduction----------------------------------------------