---
title: "Using personMatchR to match individuals across database tables"
output:
  html_document:
    df_print: paged
---

# Using personMatchR database functions to match individuals across database tables

This notebook shows how the personMatchR package can be used to match individuals across two database tables rather than within data frames held in R.

The database functions allow the database power to be leveraged, allowing much larger volumes of data to be used. For example, when developing the package, test cases across database tables containing 50 million records were producing results in under an hour.

## Prerequisites

It is recommended to have reviewed the "Using personMatchR (coffee-and-coding)" markdown document prior to this notebook to understand the basic principles of how the package works.

The database functions have been created to work with the NHSBSA DWCP and DALP databases and therefore the relevant connection strings and user accounts will be required to allow connection to either of these using the [nhsbsaR](https://github.com/nhsbsa-data-analytics/nhsbsaR) package.

## Example test case

### Install packages and libraries

In addition to the personMatchR package we will use the nhsbsaR package to connect to the database.

```{r echo = TRUE, results = "hide", message = FALSE, warning = FALSE}
# install.packages("devtools")
devtools::install_github("nhsbsa-data-analytics/personMatchR")
devtools::install_github("nhsbsa-data-analytics/nhsbsaR")
# load relevant libraries
library("dplyr")
library("dbplyr")
library("DBI")
```

### Load data and transfer to database

In absence of suitable large scale datasets we will simply transfer the test data that is included in the personMatchR package.

```{r}
# using nhsbsaR package to connect to NHSBSA DALP database
con <- nhsbsaR::con_nhsbsa(database = "DALP")

# define the user schema to be used to store test data
user_schema <- "YOUR_SCHEMA_NAME"

# copy the test datasets to the database environment as a test case example
# when creating tables specifying the field types is important
DBI::dbWriteTable(con,
                  DBI::Id(schema = user_schema, table = "PERSONMATCHR_INPUT_A"),
                  personMatchR::TEST_DF_A,
                  field.types = c(
                    ID = "number(1,0)",
                    SURNAME = "varchar(26 byte)",
                    FORENAME = "varchar(26 byte)",
                    POSTCODE = "varchar(128 byte)",
                    DOB = "date"
                  ))

DBI::dbWriteTable(con,
                  DBI::Id(schema = user_schema, table = "PERSONMATCHR_INPUT_B"),
                  personMatchR::TEST_DF_B,
                  field.types = c(
                    ID = "number(1,0)",
                    SURNAME = "varchar(26 byte)",
                    FORENAME = "varchar(26 byte)",
                    POSTCODE = "varchar(128 byte)",
                    DOB = "date"
                  ))

# establish connection to tables within the database
df_a <- con |> dplyr::tbl(from = dbplyr::in_schema(user_schema, "PERSONMATCHR_INPUT_A"))
df_b <- con |> dplyr::tbl(from = dbplyr::in_schema(user_schema, "PERSONMATCHR_INPUT_B"))

# review the tables as stored in database
df_a |> dplyr::collect()
df_b |> dplyr::collect()

```

### Format data

One of the key stages of the matching using the database functions is ensuring the data is formatted prior to calling the matching functions as there is no option to apply formatting as part of the matching function.

Please note, on large datasets the formatting functions may take a bit of time to complete execution

```{r}
# call the formatting functions
df_a <- df_a |>
  personMatchR::format_postcode_db(POSTCODE) |>
  personMatchR::format_name_db(FORENAME) |>
  personMatchR::format_name_db(SURNAME) |>
  personMatchR::format_date_db(DOB)

df_b <- df_b |>
  personMatchR::format_postcode_db(POSTCODE) |>
  personMatchR::format_name_db(FORENAME) |>
  personMatchR::format_name_db(SURNAME) |>
  personMatchR::format_date_db(DOB)

# the formatted datasets should be written back to database to allow optimal performance
df_a |> dplyr::compute(name = "PERSONMATCHR_INPUT_A_FORMAT", temporary = FALSE)
df_b |> dplyr::compute(name = "PERSONMATCHR_INPUT_B_FORMAT", temporary = FALSE)


# establish connection to formatted tables within the database
df_a_fmt <- con |> dplyr::tbl(from = dbplyr::in_schema(user_schema, "PERSONMATCHR_INPUT_A_FORMAT"))
df_b_fmt <- con |> dplyr::tbl(from = dbplyr::in_schema(user_schema, "PERSONMATCHR_INPUT_B_FORMAT"))


# review the tables as stored in database
df_a_fmt |> dplyr::collect()
df_b_fmt |> dplyr::collect()
```

### Performing matching

The matching function call is very similar to that used for matching across data frames. The key difference is that the calc_match_person_db() function call does not automatically trigger the execution. The actual processing activity will not be performed until the dplyr "compute" function is called to execute the code. This compute function will trigger the matching code and ensure the results are written to a database table where they can be accessed.

```{r}
#call the matching function
df_output <- personMatchR::calc_match_person_db(
  df_one = df_a_fmt, # first dataset
  id_one = ID, # unique id field from first dataset
  forename_one = FORENAME, # forename field from first dataset
  surname_one = SURNAME, # surname field from first dataset
  dob_one = DOB, # date of birth field from first dataset
  postcode_one = POSTCODE, # postcode field from first dataset
  df_two = df_b_fmt, # second dataset
  id_two = ID, # unique id field from second dataset
  forename_two = FORENAME, # forename field from second dataset
  surname_two = SURNAME, # surname field from second dataset
  dob_two = DOB, # date of birth field from second dataset
  postcode_two = POSTCODE, # postcode field from second dataset
  output_type = "key", # only return the key match results
  inc_no_match = TRUE # return records from first dataset without matches
)

# the match results should be written back to database before being worked with
df_output |> dplyr::compute(name = "PERSONMATCHR_OUTPUT", temporary = FALSE)

```

### Review match outputs

The match outputs can be read from the database table they have been stored in. Depending on individual use cases, this data would likely require further processing to decide how the output should be used and whether any other datasets need to be joined.

```{r}
# review the match results
con |>
  dplyr::tbl(from = dbplyr::in_schema(user_schema, "PERSONMATCHR_OUTPUT")) |>
  dplyr::collect()

# close any active database connections
DBI::dbDisconnect(con)
```
